On Sample Size for Block Bootstrap Confidence Intervals to Have Desired Coverage Rates

### 2023-08-21 Manuscript/manuscript.tex
Respond to Comments

\paragraph{Bias Corrected and Accelerated ($BC_a$) Bootstrap CI}
The $BC_a$ CI was first suggested in \citet{efron1987better}.
Let $\hat{z}_0 = \Phi^{-1}\{\#\{\hat\theta_n^{(b)} < \hat{\theta}_n\} / B\}$
for $b \in \{1, \ldots, B\}$. 
Let $Z_{(i)}$ be the original sample without the $i$th block $z_i$ for 
$i \in \{1, \ldots, k\}$, let
$\hat{\theta}_{(i)}$ be the statistic of $Z_{(i)}$,
and let $\hat{\theta}_{(.)} = \sum_{i=1}^{k} \hat{\theta}_{(i)} / k$. 
Let 
\[
\hat{a} = \frac{\sum_{i=1}^{k} (\hat{\theta}_{(.)} -
  \hat{\theta}_{(i)})^3}{6\{\sum_{i=1}^{k} (\hat{\theta}_{(.)} -
  \hat{\theta}_{(i)})^2\}^{3/2}}.
\]
Define
\[
\alpha_1 = \Phi\left(z_0 + \frac{z_{0} +
  z_{\alpha/2}}{1 - \hat{a}(\hat{z}_{0} + z_{\alpha/2})}\right)
\text{ and }
\alpha_2 = \Phi\left(z_0 + \frac{z_{0} +
  z_{1 - \alpha/2}}{1 - \hat{a}(\hat{z}_{0} + z_{1 - \alpha/2})}\right).
\]
A $(1 - \alpha)100\%~BC_a$ CI is
\[
(\hat\theta_{n, \alpha_1}^{(b)}, \quad \hat\theta_{n, \alpha_2}^{(b)}).
\]
This CI is not necessarily centered by the point estimator $\hat\theta_n$. The
$BC_a$ method corrects for bias and skewness of the $B$ bootstrap point
estimates $\{\hat\theta_n^{(1)}, \ldots, \hat\theta_n^{(B)}\}$ by including
bias-correction and acceleration factors. The acceleration factor refers to
the 
rate of change of the standard error of $\hat\theta_n$ with respect to
$\theta$.

\jy{Can we come up with a good name for the proposed method? Other people may cite this.}
\paragraph{A Proposed CI}
We propose a CI that is centered at the point estimator and uses the variation
in the bootstrap estimates to construct the error bound. This interval requires
the computation of $\bar\theta_n^{(b)}$, the mean of all bootstrap point
estimates, and $\widehat{\text{bias}} = \bar\theta_n^{(b)} - \hat\theta_n$.
A $(1 - \alpha)100\%$ CI is
\[
(\hat\theta_{n, \alpha/2}^{(b)} - \widehat{\text{bias}}, \quad
\hat\theta_{n, 1 - \alpha/2}^{(b)} - \widehat{\text{bias}}).
\]
This CI is centered around $\hat\theta_n$ and can also be rewritten
as 
\[
(\hat\theta_n + \hat\theta_{n, \alpha/2}^{(b)} - \bar\theta_n^{(b)}, \quad
\hat\theta_n + \hat\theta_{n, 1 - \alpha/2}^{(b)} - \bar\theta_n^{(b)}).
\]
It is asymmetric, as different critical values are used to compute
the lower and upper bounds.

@@

\jy{The results discussion need to be streamlined. The three paragraphs are too
  similar, and only student's ci is discussed. Questions to consider:
  1) which methods give the right coverage as sample size increases?
  2) which methods needs smaller sample sizes to get the right coverage?
  3) what's the difference between the results in positive and negative
  dependence?
  4) how does the result change as dependence gets stronger?
  }

For estimating the mean parameter~$\mu$,
Figure~\ref{fig:mu} suggests that the Student's $t$ interval is the best method.
For $\phi = -0.4$, Student's $t$ CIs over-cover 
$\mu$ for smaller $n$, whereas other CIs recover $\mu$ at the
nominal rate for all $n$ observed. For 
$\phi = -0.2$, 
all CIs recover $\mu$ at the nominal rate for all $n$. For $\phi = 0$, or when the 
sample is independently normally distributed, $n \geq 200$ seems to 
be large enough to recover $\mu$ at the nominal rate when using any CIs
 besides Student's $t$ CIs, which seem to recover $\mu$ at the nominal rate for
  all $n$.
For $\phi = 0.2$, $n \geq 800$ seems to be sufficient to recover $\mu$
at the nominal level when using
any CIs besides Student's $t$ CIS, which recover $\mu$ at the nominal
 rate for all $n \geq 200$.
For $\phi = 0.4$, $n \geq 3200$ seems to be large enough to recover $\mu$
at the nominal level when using
any CIs besides Student's $t$ CIS, which seem to recover $\mu$ at the nominal
 rate for all $n \geq 800$.

Based on Figure~\ref{fig:sigma}, Student's $t$ also seems to generally
be the best method for estimating $\sigma$.
for $\phi = -0.4$, $n \geq 1600$ is sufficient to recover $\sigma$
at the nominal level when using
any CIs besides Standard CIs, which seem to recover $\sigma$ at the nominal 
rate for all $n \geq 800$, and Student's $t$ CIs, which seem to recover $\sigma$
at the nominal rate for all $n \geq 400$. For $\phi = -0.2$, $n \geq 200$ is 
adequate to recover $\sigma$ at the nominal rate with Student $t$ CIs, 
but other CIs seem to require $n \geq 800$. For $\phi = 0$, or when the sample
is independently normally distributed, $n \geq 400$ seems to be about enough
to recover $\sigma$ for all CIs, but Student's $t$ CIs can recover $\sigma$ for
$n \geq 100$. For $\phi = 0.2$, $n \geq 800$ is adequate for CIs besides the
Student's $t$ CI, which recovers $\sigma$ well for $n \geq 200$. For 
$\phi  = 0.4$, $n \geq 1600$ seems necessary for CIs other that Student's $t$ 
CIs, which recover $\sigma$ at the nominal level for $n \geq 400$.

Based on Figure~\ref{fig:phi}, Student's $t$ again seems to generally
be the best method for estimating $\phi$. For high absolute values of $\phi$, 
Percentile, $BC$, and $BC_a$ CIs
deteriorate in coverage as $n$ increases. For $\phi = -0.4$, Standard, Student's
$t$, and Proposed CIs recover $\phi$ for $n \geq 100$. For $\phi = -0.2$, 
Standard and Proposed CIs recover $\phi$ at the nominal rate for $n \geq 200$, 
whereas this can be done for $n \geq 100$ with Student's $t$ CIs. For $\phi = 
0$, or when the sample
is independently normally distributed, Standard and Proposed CIs recover 
$\phi$ at the nominal level  for $n \geq 400$, and this is achievable for 
$n \geq 100$ with Student's $t$ CIs. For $\phi = 0.2$ and $\phi = 0.4$, Standard 
and Proposed CIs recover $\phi$ for $n \geq 200$, and Student's $t$ CIs 
recover $\phi$ for all $n \geq 100$.

To summarize, the CI type chosen should be based on the target parameter. 
When estimating $\mu$, the Student's $t$ CI seems to have the highest
 coverage
overall, although that leads to over-coverage for small sample sizes when $\phi$
 is -0.4. For a
sample with no temporal dependence, the Student's $t$ CI recover $\mu$ 
very close to the nominal rate even for small sample sizes. For positive $\phi$, a
larger sample size appears to be necessary regardless of the CI used. In the
case that $\phi$ is unknown and expected to be greater than $0$, $n > 800$
may be
necessary for block bootstrap interval estimation of $\mu$ when using the
Student's $t$ CI. If $\phi$ is known,
a lesser sample size may be sufficient. When estimating $\sigma_x$,
Student's $t$ again has the highest coverage
rates. In the case that $\phi$ is unknown and expected to be greater than $0$,
$n > 400$ may be necessary for block bootstrap interval estimation of
$\sigma_x$ when using the Student's $t$ CI. Again, if $\phi$ is known, a lesser
sample size may be enough.
When estimating $\phi$, Percentile, $BC$, and $BC_a$ CIs 
deteriorate in coverage, indicating they should not be used in this context. 
Increasing $n$ will not fix the performance of either of these intervals 
because the assumption that the coverage rate will approach the nominal level
as $n$ increases is invalid. Standard and Proposed Method CIs managed to fix
the
deterioration problem, and they perform similarly, but Student's $t$ covers
$\phi$ at the nominal rate for much smaller sample sizes. It seems that 
$n = 100$ is certainly large enough to
recover $\phi$ when using the Student's $t$ method.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{figures/norm_phi_intervals}
  \caption{100 replicate Percentile, Bias-Corrected, and Proposed Method
  Bootstrap CIs for samples of $n = 800$ for
    an AR(1) process with $\phi = 0.4$. For each replicate, the lower and upper
    bounds
    of the CIs are displayed, as well as $\hat\theta_n$ (blue) and $
    \bar\theta_n^{(b)}$ (red). }
  \label{fig:npi}
\end{figure}

@@

\jy{Keep the tex source tidy. Clean it up.}

@@

This study could be used as a guide for applied statistics courses for students
to generally understand how large of a sample size is sufficient for block
bootstrap to be used versus other inference methods. For undergraduate or
graduate students, block bootstrap is not typically a part of curriculum, but the 
results of this study can easily be used to demonstrate when it is practical to use 
this method.
This information could also prove to be useful for research using block
bootstrap
estimation of time series in domains such as econometrics. Future studies
could investigate if there are types of block bootstrap interval
construction such as $ABC$ or bootstrap-t intervals
that would more appropriately recover the parameters of a time
series. \eds{are there other methods that you know of?} 
One could also investigate the $n$ needed to make inferences about
other forms of serially dependent data such as a moving average process.

>>>>

\paragraph{Bias Corrected and Accelerated ($BC_a$) Bootstrap CI}
The $BC_a$ CI was first suggested in \citet{efron1987better}.Let
$\hat{z}_0 = \Phi^{-1}\{\#\{\hat\theta_n^{(b)} < \hat{\theta}_n\} / B\}$ for
$b \in \{1, \ldots, B\}$. Let $Z_{(i)}$ be the original sample without the
$i$th block $z_i$ for $i \in \{1, \ldots, k\}$, let $\hat{\theta}_{(i)}$ be
the statistic of $Z_{(i)}$, and let
$\hat{\theta}_{(.)} = \sum_{i=1}^{k} \hat{\theta}_{(i)} / k$. 
Let 
\[
\hat{a} = \frac{\sum_{i=1}^{k} (\hat{\theta}_{(.)} -
  \hat{\theta}_{(i)})^3}{6\{\sum_{i=1}^{k} (\hat{\theta}_{(.)} -
  \hat{\theta}_{(i)})^2\}^{3/2}}.
\]
Define
\[
\alpha_1 = \Phi\left(z_0 + \frac{z_{0} +
  z_{\alpha/2}}{1 - \hat{a}(\hat{z}_{0} + z_{\alpha/2})}\right)
\text{ and }
\alpha_2 = \Phi\left(z_0 + \frac{z_{0} +
  z_{1 - \alpha/2}}{1 - \hat{a}(\hat{z}_{0} + z_{1 - \alpha/2})}\right).
\]
A $(1 - \alpha)100\%~BC_a$ CI is
\[
(\hat\theta_{n, \alpha_1}^{(b)}, \quad \hat\theta_{n, \alpha_2}^{(b)}).
\]
This CI is not necessarily centered by the point estimator $\hat\theta_n$. The
$BC_a$ method corrects for bias and skewness of the $B$ bootstrap point
estimates $\{\hat\theta_n^{(1)}, \ldots, \hat\theta_n^{(B)}\}$ by including
bias-correction and acceleration factors. The acceleration factor refers to
the rate of change of the standard error of $\hat\theta_n$ with respect to
$\theta$.

\jy{Can we come up with a good name for the proposed method? Other people may cite this.}\mc{bias-centered? i'm not sure}
\paragraph{A Proposed CI}
We propose a CI that is centered at the point estimator and uses the variation
in the bootstrap estimates to construct the error bound. This interval requires
the computation of $\bar\theta_n^{(b)}$, the mean of all bootstrap point
estimates, and $\widehat{\text{bias}} = \bar\theta_n^{(b)} - \hat\theta_n$. A
$(1 - \alpha)100\%$ CI is
\[
(\hat\theta_{n, \alpha/2}^{(b)} - \widehat{\text{bias}}, \quad
\hat\theta_{n, 1 - \alpha/2}^{(b)} - \widehat{\text{bias}}).
\]
This CI is centered around $\hat\theta_n$ and can also be rewritten
as 
\[
(\hat\theta_n + \hat\theta_{n, \alpha/2}^{(b)} - \bar\theta_n^{(b)}, \quad
\hat\theta_n + \hat\theta_{n, 1 - \alpha/2}^{(b)} - \bar\theta_n^{(b)}).
\]
It is asymmetric, as different critical values are used to compute
the lower and upper bounds.

@@

\jy{The results discussion need to be streamlined. The three paragraphs are too
  similar, and only student's ci is discussed. Questions to consider:
  1) which methods give the right coverage as sample size increases?
  2) which methods needs smaller sample sizes to get the right coverage?
  3) what's the difference between the results in positive and negative
  dependence?
  4) how does the result change as dependence gets stronger?
  } \mc{addressed}

For estimating the mean parameter~$\mu$, Figure~\ref{fig:mu} suggests that all
methods eventually approach correct coverage of $\mu$ as sample size
increases. Student's $t$ CIs generally does not need relatively large sample
sizes to achieve correct coverage, except for samples with strong negative
dependence, in which case, they actually over-cover $\mu$ for smaller sample
sizes. For instance, for a sample with $n = 100$ and $\phi = -0.4$, the lower
bound for a Students $t$ CIs coverage of $\mu$ is greater than 95\%, whereas
the coverage intervals for other methods contain 95\%. In fact, Standard,
Percentile, $BC$, and $BC_a$, and Proposed Method CIs require similar sample
sizes to to recover $\mu$ at the nominal level for all combinations of $n$ and
$\phi$. All methods seem to require a smaller sample to recover $\mu$ at the
nominal rate when dealing with negative dependence versus positive dependence.
For example, $BC$ CIs recovers $\mu$ for $n \geq 100$ when $\phi = -0.2$, but
they only recover $\mu$ for $n \geq 800$ when $\phi = 0.2$. In addition, as a
negative dependence gets stronger, holding everything else equal, coverage
increases, which lead to the Student $t$ CI's aforementioned over-coverage. As
a positive dependence gets stronger, holding everything else equal, coverage
decreases, and a larger sample is necessary to recover $\mu$.

For estimating the standard deviation parameter~$\sigma_x$,
Figure~\ref{fig:sigma} suggests that each method can reach nominal coverage of
$\sigma_x$ if the sample is large enough, but for a given $n$ and $\phi$,
coverage of $\sigma_x$ will be lower than coverage of $\mu$ in general. Like
$\mu$, $\sigma_x$ can be covered by Student $t$ CIs with smaller sample sizes
when compared to other methods, and there is no over-coverage issue for
$\phi = -0.4$. Standard, Percentile, $BC$, $BC_a$, and Proposed Method CIs
again have similar performance. All methods seem to have slightly higher
coverage of $\sigma_x$ when $\phi$ is negative versus when $\phi$ is positive.
Regardless of the sign, coverage of $\sigma_x$ gets worse as the strength of
the temporal dependence increases.

For estimating the auto-correlation parameter~$\phi$, Figure~\ref{fig:phi}
suggests that while Standard, Student's $t$, and Proposed Method CIs do
approach correct coverage as sample size increases, Percentile, $BC$, and
$BC_a$ CIs deteriorate as sample size increases, especially as the strength of
the temporal dependence increases. Because of this, only Standard, Student's
$t$, and Proposed Method CIs should be considered as effective block bootstrap
methods to estimate $\phi$. Student's $t$ CIs once again can achieve correct
coverage with smaller sample sizes when compared to Standard and Proposed CIs
which perform similarly. Student's $t$ CIs can can recover $\phi$ at the
nominal level for $n \geq 100$ when the sample's temporal dependence is as
strong as 0.4. Coverage appears to be higher for all methods when the
dependence is negative rather than positive. Whether or not the dependence is
negative or positive, coverage of $\phi$ seems to increase slightly as the
absolute value increases.

To summarize, the CI type chosen should be based on the target parameter. When
estimating $\mu$ and $\sigma_x$, any CI will do, although Student's $t$ CIs
perform noticeably better than the others. However, when estimating $\phi$,
the choice of method is of utmost importance as to avoid coverage
deterioration. A smaller sample size is generally required to estimate a
parameter for a sample with a negative $\phi$ versus a positive $\phi$ of the
same magnitude. In order to know if coverage will increase as the strength of
the temporal dependence increases, one must know what the parameter of
interest is, and in the case of $\mu$: $\phi$'s sign.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{figures/norm_phi_intervals}
  \caption{100 replicate Percentile, Bias-Corrected, and Proposed Method
    Bootstrap CIs for samples of $n = 800$ for an AR(1) process with
    $\phi = 0.4$. For each replicate, the lower and upper bounds of the CIs
    are displayed, as well as $\hat\theta_n$ (blue) and $\bar\theta_n^{(b)}$
    (red). }
  \label{fig:npi}
\end{figure}

@@

\jy{Keep the tex source tidy. Clean it up.}\mc{addressed}

@@

This study could be used as a guide for applied statistics courses for students
to generally understand how large of a sample size is sufficient for block
bootstrap to be used versus other inference methods. For undergraduate or
graduate students, block bootstrap is not typically a part of curriculum, but
the results of this study can easily be used to demonstrate when it is
practical to use this method. This information could also prove to be useful
for research using block bootstrap estimation of time series in domains such
as econometrics. Future studies could investigate if there are types of block
bootstrap interval construction such as $ABC$ or bootstrap-t intervals, or
alternatives to block bootstrap such as AR-Sieve bootstrap that would more
appropriately recover the parameters of a time series.
\eds{are there other methods that you know of?} \mc{mentioned an alternative
  to block bootstrap}
One could also investigate the $n$ needed to make inferences about other forms
of serially dependent data such as a moving average process.
